---
title: "Homework No. 1"
author: "DATA-607 // Ben Horvath"
date: "August 28, 2018"
output:
  html_document:
    theme: null
    css: ../static/architect.css
    template: ../static/architect.html
    pandoc_args: [
      "--mathjax", "",
      "--variable", "mathjax-url:https://mathjax.rstudio.com/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML"]
  pdf_document:
    keep_tex: yes
fontsize: 11pt
geometry: null
fontfamily: mathpazo
fontfamilyoptions: osf,sc
linestretch: 1.05
header-includes:
  \usepackage{eulervm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries:

```{r, warning=FALSE}
library(RCurl)
```

First, let's load the data directly from the source:

```{r}
data_url <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.data'
original <- getURL(data_url) 
df <- read.csv(text=original, header=FALSE, stringsAsFactors=FALSE)
head(df)
```

Fill in the column names and subset to just a handful:

```{r}
colnames(df) <- c('poisonous',
                  'cap_shape', 
                  'cap_surface', 
                  'cap_color', 
                  'bruises', 
                  'odor', 
                  'gill_attachment', 
                  'gill_spacing', 
                  'gill_size', 
                  'gill_color', 
                  'stalk_shape', 
                  'stalk_root', 
                  'stalk_surface_above_ring', 
                  'stalk_surface_below_ring', 
                  'stalk_color_above_ring', 
                  'stalk_color_below_ring', 
                  'veil_type', 
                  'veil_color', 
                  'ring_number', 
                  'ring_type', 
                  'spore_print_color', 
                  'population', 
                  'habitat')

cols <- c('poisonous', 'bruises', 'gill_size', 'ring_number')
df <- df[cols]
head(df)
```

The remaining task is the de-abbreviate the data, converting each entry to a meaningful designation. 

One way to do this would be to use `gsub()`. However, a custom function might be able to make the job a little easier and a little cleaner. 

`gsub_map()` accepts a string and a mapping of pattern-replacements, performing multiple `gsub()` operations together:

```{r}
gsub_map <- function(s, mapping) {
    # Accepts a mapping of pattern-replacements on a string s, allowing more 
    # compact operations involving mutliple substitutions on the same string
    for (i in 1:length(mapping)) {
        pattern <- names(mapping[i])
        replacement <- mapping[i]
        s <- gsub(pattern, replacement, s)
    }
    return(s)
}

# Example
gsub_map('foo bar', list(foo='foo1', bar='bar1'))
```

Using `sapply()` to apply to function to each row in the columns:

```{r}
df$poisonous <- sapply(df$poisonous, gsub_map, list(e='edible', p='poisonous'))
df$bruises <- sapply(df$bruises, gsub_map, list(t='TRUE', f='FALSE'))
df$gill_size <- sapply(df$gill_size, gsub_map, list(b='broad', n='narrow'))
df$ring_number <- sapply(df$ring_number, gsub_map, list(n='0', 'o'='1', 't'='2'))

head(df)
```

Finally, convert to proper R data types where applicable:

```{r}
df$bruises <- as.logical(df$bruises)
df$ring_number <- as.integer(df$ring_number)

head(df)
sapply(df, class)
```