---
title: "DATA 607---Homework No. 3"
author: "Ben Horvath"
date: "September 16, 2018"
output:
  html_document:
    theme: null
    css: ../static/architect.css
    template: ../static/architect.html
    pandoc_args: [
      "--mathjax", "",
      "--variable", "mathjax-url:https://mathjax.rstudio.com/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML"]
  pdf_document:
    keep_tex: yes
fontsize: 11pt
geometry: null
fontfamily: mathpazo
fontfamilyoptions: osf,sc
linestretch: 1.05
header-includes:
  \usepackage{eulervm}
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries:

```{r, warning=FALSE, message=FALSE}
library(stringr)
```

# Problem 3

> Copy the introductory example. The vector `name` stores the extracted names.

```{r}
raw.data <- '555-1239Moe Szyslak(636) 555-0113Burns, C. Montgomery555-6542Rev. Timothy Lovejoy555 8904Ned Flanders636-555-3226Simpson, Homer5553642Dr. Julius Hibbert'
```

> 1. Use the tools of this chapter to rearrange the vector so that all elements conform to the standard `first_name last_name`.

First, let's try to extract just the names into a character vector where each element corresponds to a full name. This is somewhat tricky as there is no delimiter seperating names from phone numbers.

## Extracting Names

As a first attempt, let's try a simple solution, extracting all alphabetic characters:

```{r}
unlist(str_extract_all(raw.data, '[[:alpha:]]'))
```

This does indeed return every alphabetic character, but not quite the full names as we want. To rectify this, let's specify that each match must be at least 2 characters or longer:

```{r}
unlist(str_extract_all(raw.data, '[[:alpha:]]{2,}'))
```

This is much closer, but we want to keep the titles and first/last names together. Let's specify not to 'break' on spaces by including a space along with the alphabetic characters:

```{r}
unlist(str_extract_all(raw.data, '[[:alpha:] ]{2,}'))
```

An improvement, but it still 'breaking' the words on the puntuation within each name: add comma and period with the space:

```{r}
name <- unlist(str_extract_all(raw.data, '[[:alpha:],. ]{2,}'))
name
```

Perfect! `name` can be described as all sequences of characters in `raw.data` that contain alphabetic characters + period + comma + space that contain at least two or more characters.

## Standardizing `name`

Now to the core of the question. 

Normally, I would probably detect if a comma is present, and if so, split the string on the comma with `strsplit()` and `rev()` the resulting strings. 

Instead, I'll attampt to do this with RegEx and `stringr`.

Though not mentioned in the text, RegEx has a feature called matching groups, that enumerates pattern matches. This is apparent in `str_match()`:

```{r}
str_match('Simpson, Homer', '(\\w+),\\s(\\w+)')
```

Parentheses are used to break up the pattern into groups, in this case, into two words `(\\w+)`. The resulting matrix allows access to these groups via subscripts.

We can use these groups with `str_replace()` to replace the words such that 'last_name, first_name' becomes 'first_name last_name':

```{r}
str_replace('Simpson, Homer', '(\\w+),\\s(\\w+)', '\\2 \\1')
```

Mr. Burns's name is more challenging because of the first name initial, but can be solved along similar lines:

```{r}
str_replace('Burns, C. Montgomery', '(\\w+),\\s(\\w.)\\s(\\w+)', '\\2 \\3 \\1')
```

## Pulling It Altogether

Let's construct a function that converts any name string to a standardized format, and that can then be applied to `name`:

```{r}
standardize_name <- function(s) {
  last_first <- '^(\\w+),\\s(\\w+)$'
  last_first_init <-  '^(\\w+),\\s(\\w.)\\s(\\w+)$'

  s <- str_trim(s)
  
  # Test for last_name, first_name
  if (str_detect(s, last_first) == TRUE) {
    standard <- str_replace(s, last_first, '\\2 \\1')
  } 
  # Test for last_name, first_initial middle_name
  else if (str_detect(s, last_first_init) == TRUE) {
    standard <- str_replace(s, last_first_init, '\\2 \\3 \\1')
  }
  else {
    standard <- s
  }
  return(standard)
}
```

Test to make sure the function works as desired:

```{r}
standardize_name('Simpson, Homer')
standardize_name('Burns, C. Montgomery')
standardize_name('Lisa Simpson')
```

Finally---apply to entire `name` vector:

```{r}
lapply(name, FUN=standardize_name)
```



> 2. Construct a logical vector indicating whether a character has a title (i.e., _Rev._ and _Dr._).

Using `str_detect()` and the pipe operator:

```{r}
str_detect(name, 'Dr.|Rev.')
```



> 3. Construct a logical vector indicating whether a character has a second name.

Mr. Burns is the only one with a second name in this list, and it is observable as a single capital alphabetic character and a period, followed by a space:

```{r}
str_detect(name, '[A-Z].\\s')
```



# Problem 4

> Describe the types of strings that conform to the following regular expressions and construct an example that is matched by the regular expression.

> 1. `[0-9]+\\$`

This matches a sequence of numbers (of any size) that ends with a dollar sign (note the double backslash before $).

```{r}
str_extract_all('1$', '[0-9]+\\$')
str_extract_all('1000000$ is a lot of money', '[0-9]+\\$')
```

> 2. `\\b[a-z]{1,4}\\b`

This command will return every sequence that starts with a lowercase letter, is composed of lowercase letters, and is between 1 and 4 characters in length. The command below should only return _test_ and _for_ as they are the only words that meet all the criteria.

```{r}
str_extract_all('test test1 amalgamation for Ben eB4n', '\\b[a-z]{1,4}\\b')
```

> 3. `.*?\\.txt$`

This matches strings resembling files with the extension `.txt`. The first two commands will return a match and the third will not:

```{r}
str_extract('notes.txt', '.*?\\.txt$')
str_extract('KKK*8293ufskf.txt', '.*?\\.txt$')
str_extract('txt.exe', '.*?\\.txt$')
```

> 4. `\\d{2}/\\d{2}/\\d{4}`

This is a date format, like '09/08/2018' for September 8, 2018, or (perhaps if European) for August 9, 2018. The first should return a match, the second will not.

```{r}
str_extract('09/08/2018', '\\d{2}/\\d{2}/\\d{4}')
str_extract('2018/09/08', '\\d{2}/\\d{2}/\\d{4}')
```

> 5. `<(.+?)>.+?</\\1>`

This looks like it would match HTML or similar style tags (without attributes like `href`), like:

```{r}
str_extract('<h1>Trump Shocks Nation</h1>', '<(.+?)>.+?</\\1>')
str_extract('<p>The president\'s decline continues this week as...</p>', '<(.+?)>.+?</\\1>')
str_extract('<a href="google.com">Google</a>', '<(.+?)>.+?</\\1>')
```



# Problem 9

> The following code hides a secret message. Crack it with `R` and regular expressions.

Try out different combinations of character classes until one works:

```{r}
msg <- 'clcopCow1zmstc0d87wnkig7OvdicpNuggvhryn92Gjuwczi8hqrfpRxs5Aj5dwpn0TanwoUwisdij7Lj8kpf03AT5Idr3coc0bt7yczjatOaootj55t3Nj3ne6c4Sfek.r1w1YwwojigOd6vrfUrbz2.2bkAnbhzgv4R9i05zEcrop.wAgnb.SqoU65fPa1otfb7wEm24k6t3sR9zqe5fy89n6Nd5t9kc4fE905gmc4Rgxo5nhDk!gr'
str_c(unlist(str_extract_all(msg, '[[:upper:].]')), collapse='')
```

:)