---
title: "DATA 607---Homework No. 7"
author: "Ben Horvath"
date: "October 14, 2018"
output:
  html_document:
    theme: null
    css: ../static/architect.css
    template: ../static/architect.html
    pandoc_args: [
      "--mathjax", "",
      "--variable", "mathjax-url:https://mathjax.rstudio.com/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML"]
  pdf_document:
    keep_tex: yes
fontsize: 11pt
geometry: null
fontfamily: mathpazo
fontfamilyoptions: osf,sc
linestretch: 1.05
header-includes:
  \usepackage{eulervm}
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load general libraries:

```{r, warning=FALSE, message=FALSE}
library(dplyr)
# install.packages('jsonlite', 'rvest')
```

This document explores loading JSON, XML, and HTML table representations of the same data into R data frames. The data itself is three records, each containing information on popular young adults books.

For your understanding, this is what the JSON data looks like:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
readLines('books.json')
```

See the Remarks section toward the end of the document.



# JSON

```{r, warning=FALSE, message=FALSE}
library(jsonlite)
from_json <- fromJSON('books.json', flatten=TRUE)
head(from_json)
```



# XML

```{r, warning=FALSE, message=FALSE}
library(XML)
xml_doc <- xmlParse('books.xml')
root <- xmlRoot(xml_doc)
data <- xmlSApply(root, function(x) xmlSApply(x, xmlValue))
from_xml <- data.frame(t(data), row.names=NULL) %>%
    select(colnames(from_json))
head(from_xml)
```

Note the `authors` column, see Remarks below.



# HTML

Using the `rvest` package by our dear Hadley Wickham:

```{r, warning=FALSE, message=FALSE}
library(rvest)
html_doc <- read_html('books.html')
html_tbl <-  html_doc %>% 
    html_nodes('table') %>% 
    html_table
from_html <- as.data.frame(html_tbl[1])
head(from_html)
```



# Remarks

These three data frames are almost identical. The author data in the XML-derived data frame is printed as one author, `Stephenie MeyerIllustrator Jones` which could become a problem for some analyses of this data. In the othre two data frames, it is appropriately represented with a seperator character, `Stephenie Meyer, Illustrator Jones`.

In the data recording stage, it would probably be a better design decision to put quotation marks around each authors' name, and use a semicolon (rather than comma) to seperate them. This would reduce problems trying to seperate out the individual authors in the author column of the R data frame.

A final observation: The `XML` library converts much of the data to factors, where `rvest` does a good job using the more appropriate integer, numeric, and character data types.
